1. [Launch Terra on AnVIL](https://anvil.terra.bio/#workspaces)

1. In the drop-down menu on the left, navigate to "Workspaces". Click the triple bar in the top left corner to access the menu. Click "Workspaces".

    ```{r, echo=FALSE, fig.alt='Screenshot of Terra drop-down menu. The "hamburger" button to extend the drop-down menu is highlighted, and the menu item "Workspaces" is highlighted.'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BLTCaogA04bbeSD1tR1Wt-mVceQA6FHXa8FmFzIARrg/edit#slide=id.g11f12bc99af_0_2")
    ```

1. Click on the **plus icon** near the top of left of the page.

    ```{r, echo=FALSE, fig.alt='Screenshot of Terra Workspaces page. The "+" is highlighted.'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BLTCaogA04bbeSD1tR1Wt-mVceQA6FHXa8FmFzIARrg/edit#slide=id.g11e8aba06f7_0_6")
    ```

1. Name your Workspace and select the Billing Project as indicated by your instructor.

    ```{r, echo=FALSE, fig.alt='Screenshot of Terra dialog box for creating a new Workspace. The text box labeled "Workspace name" and the drop-down menu labeled "Billing project" are highlighted.'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BLTCaogA04bbeSD1tR1Wt-mVceQA6FHXa8FmFzIARrg/edit#slide=id.g11f12bc99af_0_18")
    ```

1. Do not change the Bucket Location or Authorization Domain. You can add a Description if you wish.

1. Click "CREATE WORKSPACE". The new Workspace should now show up under your Workspaces.

    ```{r, echo=FALSE, fig.alt='Screenshot of Terra dialog box for creating a new Workspace. The "Create Workspace" button is highlighted.'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1BLTCaogA04bbeSD1tR1Wt-mVceQA6FHXa8FmFzIARrg/edit#slide=id.g11f12bc99af_0_30")
    ```
